Starcode tutorial

### 2015-11-17 doc/tutorial.tex
Updated the tutorial.

@@

@@

The previous example is a standard run of Starcode on sequence
data. By default, Starcode is verbose and gives some information on
how the run is performed before it outputs the clusters. 
In this case, we learn that Starcode used a single CPU, that
the input file \texttt{test/test\char`_file.txt} is in raw format
(\textit{i.e.} plain text), and that the threshold distance for
similarity search was set to 2. The output propoer starts after
the line \texttt{progress: 100.00\%}. It consists of five lines
(one per cluster) where the sequence is the centroid of the cluster
(the irepresentative sequence) and the number is its size (the
number of sequences in the cluster).

@@

>>>>

\begin{verbatim}
./starcode
\end{verbatim}

@@

\begin{verbatim}
Usage:  starcode [options]

  general options:
    -d --dist: maximum Levenshtein distance (default auto)
    -t --threads: number of concurrent threads (default 1)
    -r --cluster-ratio: minumum cluster size ratio (default 5)
    -s --sphere: sphere clustering (default message passing)
    -q --quiet: quiet output (default verbose)
    -v --version: display version and exit

  input/output options (single file, default)
    -i --input: input file (default stdin)
    -o --output: output file (default stdout)

  input options (paired-end fastq files)
    -1 --input1: input file 1
    -2 --input2: input file 2

  output format options
       --print-clusters: outputs cluster compositions
       --non-redundant: remove redundant sequences
       --seq-id: also print sequence id numbers (1-based)
\end{verbatim}

@@

Let us get started with a simple example. At the command line,
issue the command below.

In principle, you should obtain the following output.

By default, Starcode is verbose and gives some information on
how the run is performed before it outputs the clusters. 
In this case, we learn that Starcode used a single CPU, that
the input file \texttt{test/test\char`_file.txt} is in raw format
(\textit{i.e.} plain text), and that the threshold distance for
similarity search was set to 2. The output propoer starts after
the line \texttt{progress: 100.00\%}. It consists of five lines
(one per cluster) where the sequence is the centroid of the cluster
(the irepresentative sequence) and the number is its size (the
number of sequences in the cluster).

@@

\section{Clustering options}

Coming soon...

\section{Documenting clusters}

Starcode also allows you to track the sequences that end up in a
given cluster. Here, `sequences' can be considered as either input
sequences, or distinct nucleotide sequences. The option
\texttt{--print-clusters} shows the distinct nucleotide sequences
that constitute the cluster. Below is and example.

\begin{verbatim}
./starcode -q --print-clusters test/test_file.txt
AGGGCTTACAAGTATAGGCC  7 AGGGCTTACAAGTATAGGCC,AGGGGTTACAAGTCTAGGCC
CCTCATTATTTGTCGCAATG  7 CCTCATTATTTACCGCAATG,CCTCATTATTTGTCGCAATG
GGGAGCCCACAGTAAGCGAA  7 GGAAGCCCACAGCAAGCGAA,GGGAGCCCACAGTAAGCGAA
TAGCCTGGTGCGACTGTCAT  7 TAACCTGGTGCGACTGTTAT,TAGCCTGGTGCGACTGTCAT
TGCGCCAAGTACGATTTCCG  7 TGCGCCAAGTAAGAATTCCG,TGCGCCAAGTACGATTTCCG
\end{verbatim}

With this type of output, we know that the first cluster
consists of two distinct sequences, namely the centroid itself
and \texttt{AGGGGTTACAAGTCTAGGCC}, which together amount to 7
sequences. This kind of output is useful when you want to convert
sequences to their centroid.

As mentioned above, `sequences' can also be viewed as input sequences
in which case they represent different entities, even when their
nucleotide sequences are the same. These sequences are typically
different reads from a sequencing experiment. The option
\texttt{--seq-id} prints the ID numbers of the input sequences that
end up in each cluster. Here is an example.

\begin{verbatim}
./starcode -q --seq-id test/test_file.txt
AGGGCTTACAAGTATAGGCC  7 1,4,5,13,17,20,22
CCTCATTATTTGTCGCAATG  7 3,14,19,27,30,31,32
GGGAGCCCACAGTAAGCGAA  7 6,7,10,12,15,24,25
TAGCCTGGTGCGACTGTCAT  7 8,9,16,21,28,33,35
TGCGCCAAGTACGATTTCCG  7 2,11,18,23,26,29,34
\end{verbatim}

With this type of output, we know that the first cluster consists of
input sequences number 1, 4, 5, 13, 17,20 and 22. This kind of input
is useful when sequences have different origins and you want to
know the composition of a cluster relative to the different sources.
Note that the first input sequence is given ID number 1 and not 0.

Note that options \texttt{--print-clusters} and \texttt{--seq-id} are
not mutually exclusive, and they may be used together. These options
have only mild effects on the speed and the memory footprint of
Starcode.

\section{Input formats}

Starcode accepts FASTA files, FASTQ files and plain text files.
The directory \texttt{test} contains some example files. We have
already seen what happens on plain text files in the previous
examples. For FASTA files, Starcode is run as before. The FASTA
headers are simply ignored.

\begin{verbatim}
./starcode -q test/test_file.fasta
AGGGCTTACAAGTATAGGCC  3
CCTCATTATTTGTCGCAATG  1
TGCGCCAAGTACGATTTCCG  1
\end{verbatim}

FASTQ files present a complication because they can be paired.
In the case of a single FASTQ file, Starcode is run as before.
The FASTQ headers and the quality are ignored.

\begin{verbatim}
./starcode -q test/test_file1.fastq
AGGGCTTACAAGTATAGGCC  3
CCTCATTATTTGTCGCAATG  1
TGCGCCAAGTACGATTTCCG  1
\end{verbatim}

When there are two paired FASTQ files, you must tell Starcode to
expect two input files. The first way is to use the \texttt{-1}
flag for the first file and the \texttt{-2} for the second.
For more clarity, you can use the longer form \texttt{--input1}
and \texttt{input2} instead. Below is an example.

\begin{verbatim}
./starcode -q -1 test/test_file1.fastq -2 test/test_file2.fastq
AGGGCTTACAAGTATAGGCC/AAGGGCTTACAAGTATAGGC 3
CCTCATTATTTGTCGCAATG/ACCTCATTATTTGTCGCAAT 1
TGCGCCAAGTACGATTTCCG/ATGCGCCAAGTACGATTTCC 1
\end{verbatim}

As you can see, the output also changes. The two paired sequences
are lumped next to each other and each centroid now consists of
a pair of sequences. It is important to mention that in this
mode, the edit distance between two pairs of sequences is the sum
of the pairwise edit distances. For instance, if the edit distance
is set to 3 with \texttt{-d3}, the pair \texttt{AAA/TTT} is a
neighbor of the pair \texttt{CAA/GGT}, but not of the pair
\texttt{CCC/GGG}.

\section{Getting a non redundant sample}

By default, Starcode clusters the input sequences and outputs the
clusters. In some application, it is interesting to obtain a non
redundant sample of the input. In general, it is easy to do so by
collecting the centroid sequences, but because Starcode discards
extra information from FASTA and FASTQ files, we provide the
option \texttt{--non-redundant} for cases that the format has to
be respected.

\begin{verbatim}
./starcode -q --non-redundant test/test_file1.fastq
@seq1/1
AGGGCTTACAAGTATAGGCC
+
BBBBBBBBBBBBBBBBBBBB
@seq3/1
CCTCATTATTTGTCGCAATG
+
BBBBBBBBBBBBBBBBBBBB
@seq2/1
TGCGCCAAGTACGATTTCCG
+
BBBBBBBBBBBBBBBBBBBB
\end{verbatim}

When two FASTQ files are used with this option, nothing is
printed on the screen. Instead, two new FASTQ files are created.
Note that the sequences are still sorted in alphabetical order,
so their order will usually not be the same as the original input.
